\mbox{Specifying Imperative Higher-Order Programs}

### 2009-10-02 papers/esop2010/main.tex
go

\newcommand{\wpre}{{\bf wp}}
\newcommand{\sep}{ ~|~ }
\newcommand{\letml}{{\bf let}}
\newcommand{\inml}{{\bf in}}
\newcommand{\ifml}{{\bf if}}
\newcommand{\thenml}{{\bf then}}
\newcommand{\elseml}{{\bf else}}
\newcommand{\refml}{{\bf ref}}
\newcommand{\propml}{{\bf prop}}
\newcommand{\boolml}{{\bf bool}}
\newcommand{\recml}{{\bf rec}}
\newcommand{\efft}[1]{\langle #1 \rangle}
\newcommand{\lift}[1]{\langle #1 \rangle}
\newcommand{\alist}[1]{\overline{#1} }
\newcommand{\Trueml}{{\bf True}}
\newcommand{\Falseml}{{\bf False}}
\newcommand{\trueml}{{\bf true}}
\newcommand{\falseml}{{\bf false}}
\newcommand{\unitml}{{\bf unit}}
\newcommand{\coq}{Coq}
\newcommand{\who}{Who}
\newcommand{\correct}[1]{c(#1)}

@@

\begin{document}
\title{\mbox{Specifying Imperative Higher Order Programs}}

\author{Johannes Kanig \and Jean-Christophe Fill\^atre}
\institute{
Laboratoire de Recherche en Informatique, Univ Paris-Sud, CNRS, Orsay F-91405\\
INRIA Futurs, ProVal, Parc Orsay Universit\'e, F-91893
}
%TODO affiliation
%TODO find notions for:
%  * ``semantic values'' v
%  * ``store types'' <ε>
%

@@

% ICFP
%  We present a weakest precondition calculus for an imperative,
%  higher-order programming language, restricted to non-aliasing global
%  references. The main feature of the calculus is {\em effect
%    polymorphism}, {\em i.e.}, higher-order functions can be specified by
%  generalizing over the effects and specifications of their functional
%  arguments. This allows truly modular reasoning about effectful
%  higher-order functions. We achieve this by defining an extension of
%  a standard higher-order logic, more suitable to reason about
%  modifications of the store. The obtained proof obligations, however,
%  are expressible in the standard logic and can be proved using
%  standard theorem provers. We show that the necessary annotations
%  and the obtained proof obligations are quite natural.
%

% Workshop ML
%  We present \who, a tool for verifying effectful higher-order
%  functions. It features {\em Effect polymorphism}, higher-order logic
%  and the possibility to reason about state in the logic, which enable
%  highly modular specifications of generic code. Several small
%  examples and a larger case study demonstrate its usefulness. The
%  \who~tool is intended to be used as an intermediate language for
%  verification tools targeting ML-like programming languages.
%

\begin{abstract} We present a specification language suited for reasoning
  about effectful higher-order programs. The language's main features are
  higher-order logic and effect polymorphism, both of which enable a truly
  modular treatment of higher-order functions. To our knowledge, this is the
  first time that effect polymorphism is put to use in program verification.
  We also define a weakest precondition calculus for this language which
  generates proof obligations in higher-order logic. A simple mechanism based
  on regions permits to restrict the analysis to alias-free programs, thus
  obtaining much simpler proof obligations. Several examples demonstrate the
  practicality of the system. A prototype implementation exists and has been
  used to prove the correctness of these and more examples.  
\end{abstract}

\section{Introduction}
%TODO Hoare logic = sharing between logic and prog language + alias free

also one or two examples in the described language

@@

\section{A programming language for specification}

technical presentation:
\begin{itemize}
  \item ANF
  \item values defined by typing 
  \item  two different types of arrows
  \item  distinction bool and prop
  \item  notion of substitution - correspondence χ κ
  \item semantics is standard (goes to appendix )
\end{itemize}

@@

\paragraph{Program typing.}
The typing of programs, written $Γ|- t : τ,ε$, assigns a type $τ$ and an
effect $ε$ to an expression $t$, given an environment. This typing is
presented in {\em A-normal form}, a normal form where all intermediate values
are \letml-introduced. All values are also programs with no effect (rule {\sc
Value}). An application must consist of two values and the effect of the
overall expression is precisely the effect of the called function (rule {\sc
App}). There are two rules for \letml, due to the value restriction. One only
can generalize type, effect and region variables when the bound expression is
a value. In this case (rule {\sc LetV}), only the second term may produe
effects. In the other case (rule {\sc Let}), no generalization takes place,
but both terms can have effects. The {\sc Let} rule actually obliges both
terms to have the same effect, but together with the {\sc Sub} rule, which can
increase the effect of an expression, this can always be achieved.
Finally, \ifml-expressions are typed in the usual way, both branches have to
have the same effect.

@@

The reader may have wondered why the specification language does not contain
any logical connectives. These can of course be declared independently, for
example the conjuction is just a function: \begin{equation*} /\ :  \propml ->
  \propml -> \propml \end{equation*} Actually, when equality is given, all
  logical operations (even quantification) can be defined in the logic.
%TODO citation needed
In the remainder of the paper, we suppose predefined the usual logical
connectives conjunction, disjunction, implication and equivalence, as well as
universal and existential quantification. We also assume all the arithmetic
operations as well as the constructor $(t,t)$  and the accessors $fst$ and
$snd$ for pairs.

% TODO TODO
%Finally, we introduce two symbols
%\begin{eqnarray*}
%  \oplus &:&∀ e1 e2 . \efft{ e1 } -> \efft{ e2 } -> \efft { e1  e2 }\\
%  |_{ε} &:& ∀ e. \efft{ e ε } -> \efft{ ε }.
%  !! & : & ∀αρ e. α~\refml_ρ -> \efft{ ρ e } -> α
%\end{eqnarray*}
%The first one denotes a combining operation for stores; the resulting store
%contains all the mappings of the second store, as well as all the mappings of
%the first one, as long as they are not present in the second one.
%union of the two arguments, in the sense that its domain is the union of its
%arguments.

@@

A way out is to restrict programs to alias free programs. Fortunately, it is
easy to modify the presented typing rules to render all programs with aliasing
ill-typed (see Fig~\ref{fig:aliasing}). In the following, we will use regions
to identify memory locations entirely; each region shall describe precisely
one memory location. Starting from this assumption, a program can actually
exhibit {\em two} forms of aliasing. In the first one, it may use several
names (region variables) for the same memory location. This can happen when a
function is generalized over, say, two region variables, assuming they are
different, and is then instantiated with the same region variable twice. The
classical example is the following program
%TODO

which is only correct if $ ρ1 $ and $ ρ2 $ are indeed different. To enforce
this, we add a side condition to the instantiation: the substitution $σ$ must
be {\em compatible} with the type $τ$. To understand what compatibility means,
consider an {\em atomic} substitution $[ρ|->ρ']$. It is compatible with any
type which does not contain $ρ'$. Similarly, the substitution $[e|-> ρ1  ρ2
e'$] is compatible with any type which does not contain $ ρ1 , ρ2 $ nor $e'$.
Expressed more generally, an atomic substitution is compatible with a type if
the type does not contain any region or effect variables which are in the
range of the substitution. Finally, a composite substitution is compatible
%TODO atomic and composite
with a type if each atomic substitution is compatible with the result of the
application of the previous substitution. In this way, it becomes impossible
to have different names for a single reference cell.

@@

\section{A Simple Weakest Preconditions Calculus}

@@

\section{Extensions}

subtyping: effects

\section{Implementation}

%what has been implemented

@@

%details
% lines of code : less than 3000 lines of ocaml
% limited form of type and effect inference
%

@@

\begin{center}
    \begin{tabular}{ | r | r | r | r |}
    \hline
    Program & Code & Spec & Proofs \\ \hline
    Array &  &  &  \\
    Hashtbl & & &  \\ \hline
    Memo/Ymemo & & &  \\ \hline
    Koda-Ruskey~\cite{KanigFilliatre09wml} & & &  \\ \hline
    \end{tabular}
\end{center}

\subsection{memo}
\subsection{ymemo}

\section{Related Work}

\section{Conclusion}

@@

We are currently investigating the possibility to obtain first-order proof
obligations, to be able to use SMT solvers~\cite{RanTin-SMTLIB}, which would
avoid manual proofs in many cases. An obvious solution would be to reuse
existing encodings of higher-order logic in first-order logic, such as the one
by Pottier and Gauthier~\cite{pottier-gauthier-hosc}. This technique has been
successfully employed by the Pangolin system~\cite{regis-gianas-pottier-08}.
Another option comes from our observation that in many cases, the proof
obligations for higher-order functions can actually be expressed in
first-order logic, when one replaces top-level {\em quantifications} of
functional arguments and invariants by top-level {\em declarations}. Doing
this translation manually, we have obtained first-order proof
obligations for the {\tt Array.iter} function which are all discharged by
SMT solvers. We don't know if such a transformation can always be carried out.

what about correspondence between $τ ->^\emptyset τ$ and $τ -> τ$?

>>>>

\newcommand{\wpre}{{\bf wp}}
\newcommand{\sep}{ ~|~ }
\newcommand{\letml}{{\bf let}}
\newcommand{\inml}{{\bf in}}
\newcommand{\ifml}{{\bf if}}
\newcommand{\thenml}{{\bf then}}
\newcommand{\elseml}{{\bf else}}
\newcommand{\refml}{{\bf ref}}
\newcommand{\propml}{{\bf prop}}
\newcommand{\boolml}{{\bf bool}}
\newcommand{\recml}{{\bf rec}}
\newcommand{\efft}[1]{\langle #1 \rangle}
\newcommand{\lift}[1]{\langle #1 \rangle}
\newcommand{\alist}[1]{\overline{#1} }
\newcommand{\Trueml}{{\bf True}}
\newcommand{\Falseml}{{\bf False}}
\newcommand{\trueml}{{\bf true}}
\newcommand{\falseml}{{\bf false}}
\newcommand{\unitml}{{\bf unit}}
\newcommand{\coq}{Coq}
\newcommand{\who}{Who}
\newcommand{\ocaml}{Ocaml}
\newcommand{\correct}[1]{c(#1)}

@@

\begin{document}
\title{\mbox{Specifying Imperative Higher-Order Programs}}

\author{Johannes Kanig \and Jean-Christophe Fill\^atre}
\institute{
Laboratoire de Recherche en Informatique, Univ Paris-Sud, CNRS, Orsay F-91405\\
INRIA Futurs, ProVal, Parc Orsay Universit\'e, F-91893
}
%TODO affiliation
%TODO explain somewhere the correspondence store type <-> map, effect variable
%<-> domain, region <-> key
%TODO the constants dealing with maps

%external TODO : proofs for Hashtbl
%external TODO : proofs of system
%

@@

\begin{abstract} We present a specification language suited for reasoning
  about effectful higher-order programs. The language's main features are
  higher-order logic and effect polymorphism, both of which enable a truly
  modular treatment of higher-order functions.  We also define a weakest
  precondition calculus for this language which generates proof obligations in
  higher-order logic.  To our knowledge, this is the first time that effect
  polymorphism is put to use in program verification.  A simple mechanism
  based on regions permits to restrict the analysis to alias-free programs,
  thus obtaining much simpler proof obligations. Several examples demonstrate
  the practicality of the system. A prototype implementation exists and has
  been used to prove the correctness of these and more examples.
\end{abstract}

\section{Introduction}

The purpose of the present paper is to introduce a simple weakest precondition
calculus for imperative higher-order programs. Higher-order functions are a
key feature, not only of purely functional languages like Haskell and the pure
fragment of ML, but also in combination with imperative features, like
references. However, reasoning about higher-order functions is difficult, and
especially so in the presence of effectful computations. Any function's
specification depends on its arguments. In a higher-order setting, arguments
may be functions that have their own specification. In consequence, to be
modular, the specification of a higher-order function can only be formulated
{\em depending on the specifications of its functional arguments}. In a
setting where side effects are possible, one also needs to formulate the
specification {\em depending on the side effects of the functional arguments}.
We claim that to be truly modular, the specification of a higher-order
function must be generic with respect to the specification {\em and} the
effect of its arguments.

In this paper, we first present a specification language combining features of
a programming language and logic. It uses an effect system and a region
calculus to track references. The effect system contains {\em effect
polymorphism}, thus dealing well with higher-order functions. The
specifications can be expressed in higher-order logic, to be able to state
properties of functional arguments. We show how two very simple modifications
to the system let us obtain alias-free programs, with the aim of obtaining
much simpler proof obligations.

We then proceed to define a weakest precondition calculus based on this
language, transforming an annotated program into a logic formula, which has to
be valid for the program to be correct. This calculus is very straight-forward
and syntax-directed.

Finally, we show the usefulness of the system by giving several examples of
annotated programs. All the programs have been proved correct using our
prototype implementation called \who.

\paragraph{Outline.}
Section~\ref{sec:genpic} gives an introduction to the described language as
well as an example. Sections~\ref{sec:syntax} and \ref{sec:wp} give the
details about the programming language and the weakest precondition calculus.
In section~\ref{sec:implem} we speak about the implementation and show some
examples. Finally, section~\ref{sec:related} discusses related work.

\section{The general picture and a motivating example}
\label{sec:genpic}
%TODO Hoare logic = sharing between logic and prog language + alias free
also one or two examples in the described language

@@

\section{A programming language for specification}
\label{sec:syntax}

technical presentation:
\begin{itemize}
  \item  notion of substitution - correspondence χ κ
  \item semantics is standard (goes to appendix )
\end{itemize}

@@

\paragraph{Program typing.}
The typing of programs, written $Γ|- t : τ,ε$, assigns a type $τ$ and an
effect $ε$ to an expression $t$, given an environment. This typing is
presented in {\em A-normal form}, a normal form where all intermediate values
are \letml-introduced. This is not a restriction, as any program can be
transformed an A-normal form.

All values are also programs with no effect (rule {\sc
Value}). An application must consist of two values and the effect of the
overall expression is precisely the effect of the called function (rule {\sc
App}). There are two rules for \letml, due to the value restriction. One only
can generalize type, effect and region variables when the bound expression is
a value. In this case (rule {\sc LetV}), only the second term may produe
effects. In the other case (rule {\sc Let}), no generalization takes place,
but both terms can have effects. The {\sc Let} rule actually obliges both
terms to have the same effect, but together with the {\sc Sub} rule, which can
increase the effect of an expression, this can always be achieved.
Finally, \ifml-expressions are typed in the usual way, both branches have to
have the same effect.

@@

The reader may have wondered why the specification language does not contain
any logical connectives. These can of course be declared independently, for
example the conjuction is just a function: \begin{equation*} /\ :  \propml ->
\propml -> \propml \end{equation*} Actually, when equality is given, all
logical operations (even quantification) can be defined in the
logic~\cite{andrews86}.  In the remainder of the paper, we suppose
predefined the usual logical connectives conjunction, disjunction,
implication and equivalence, as well as universal and existential
quantification. We also assume all the arithmetic operations as well as the
constructor $(t,t)$  and the accessors $fst$ and $snd$ for pairs.

%TODO pre and post
% TODO TODO
%Finally, we introduce two symbols
%\begin{eqnarray*}
%  \oplus &:&∀ e1 e2 . \efft{ e1 } -> \efft{ e2 } -> \efft { e1  e2 }\\
%  |_{ε} &:& ∀ e. \efft{ e ε } -> \efft{ ε }.
%  !! & : & ∀αρ e. α~\refml_ρ -> \efft{ ρ e } -> α
%\end{eqnarray*}
%The first one denotes a combining operation for stores; the resulting store
%contains all the mappings of the second store, as well as all the mappings of
%the first one, as long as they are not present in the second one.
%union of the two arguments, in the sense that its domain is the union of its
%arguments.

@@

A way out is to restrict programs to alias free programs. Fortunately, it is
easy to modify the presented typing rules to render all programs with aliasing
ill-typed (see Fig~\ref{fig:aliasing}). In the following, we will use regions
to identify memory locations entirely; each region shall describe precisely
one memory location. Starting from this assumption, a program can actually
exhibit {\em two} forms of aliasing. In the first one, it may use several
names (region variables) for the same memory location. This can happen when a
function is generalized over, say, two region variables, assuming they are
different, and is then instantiated with the same region variable twice. The
classical example is the following program
%TODO
\begin{who}
  let f [ρ ρ'] (x : int ref(ρ)) (y : int ref(ρ')) =
  { }
  x := !x + 1;
  !y
  { r : r = !!y old }
\end{who}

which is only correct if $ ρ $ and $ ρ' $ are indeed different. To enforce
this, we add a side condition to the instantiation: the substitution $σ$ must
be {\em compatible} with the type $τ$. To understand what compatibility means,
consider an {\em atomic} substitution $[ρ|->ρ']$. It is compatible with any
type which does not contain $ρ'$. Similarly, the substitution $[e|-> ρ1  ρ2
e'$] is compatible with any type which does not contain $ ρ1 , ρ2 $ nor $e'$.
Expressed more generally, an atomic substitution is compatible with a type if
the type does not contain any region or effect variables which are in the
range of the substitution. Finally, a composite substitution is compatible
%TODO atomic and composite
with a type if each atomic substitution is compatible with the result of the
application of the previous substitution. In this way, it becomes impossible
to have different names for a single reference cell.

@@

\section{A Simple Weakest Preconditions Calculus}
\label{sec:wp}

@@

%TODO lemmas about correctness
\section{Implementation}
\label{sec:implem}

@@

The implementation consists of less than 3000 lines of \ocaml\ code, including
a limited form of type and effect inference.

@@

In this section, we will show two simple examples which use higher-order
features.
\subsection{A simple memoization function}

\begin{figure}[tpb]
\begin{who}
type ('a,'b) map
logic mem : forall ('a, 'b) . 'a -> ('a,'b) map -> bool
logic get : forall ('a, 'b) . 'a -> ('a,'b) map -> 'b
logic set : forall ('a, 'b) . 'a -> 'b -> ('a,'b) map -> ('a,'b) map

let stores f0 m =
  forall x : Z, mem x m = true -> get x m = f0 x

parameter table : (int,int) map ref(ρ)

let memo [e] (f0 : int -> int) (f : int ->{} int) (x : int) = 
  {pre f x /\ stores f0 !!table}
  if mem x !table then get x !table
  else
    let z = f x in
    table := set x z !table;
    z
  {r : post f x r}
\end{who}
  \caption{The code along with the specification of the {\tt memo} function}
  \label{fig:memo}
\end{figure}

Our first example is a simple memoization function {\tt memo} for (pure)
functions.  It takes as an argument a function (from {\tt int} to {\tt int}
here) and an integer and returns the same result as the functional argument
would; however, before executing the argument, it will look up the integer in
a table to see if the function hasn't been called yet on this integer. If so,
the table's value is returned, otherwise we call the function, store the
result in the table, and return it. Fig~\ref{fig:memo} shows the code along
with the specification. The specification is very simple; it states that {\tt
memo} will return a value that verifies the postcondition of {\tt f}, as long
is initially we have the precondition of f and as long as the table stores
only valid key/value pairs. As we are not interested in the exact way of the
implemenation of the map, we simply model it here by a reference to a
functional map with get/set functions.

\subsection{ymemo}

\begin{figure}[tpb]
\begin{who}
  let realizes 
  let ymemo (f0 : int -> int) 
            (ff : (int ->{table} int) ->{} int ->{table} int) = 
  { forall (k : [[int ->{table} int ]]),
      { realizes f0 k} ff k { r : realizes f0 r} }
  let rec f (x : int) : int ->{table} int =
    { stores f0 !!table }
    if mem x !table then get x !table
    else
      let z = ff f x in
      table := set x z !table; z
    {r : r = f0 x /\ stores f0 !!table}
  in f
  { rf : realizes f0 rf }
\end{who}
\caption{The {\tt ymemo function} and its specifacation}
  \label{fig:<+label+>}
\end{figure}<++>

\subsection{All examples - an overview}
\begin{figure}[tbp]
\begin{center}
    \begin{tabular}{ | r | r | r | r |}
    \hline
    Program & Ocaml Code & Code + Spec in \who & Proofs \\ \hline
    Memo/ymemo & 15 loc & 26 loc & 90 lop  \\ \hline
    {\tt Array} & 69 loc & 170 loc  & 320 lop  \\ \hline
    {\tt Hashtbl} (incomplete) & & &  \\ \hline
    Koda-Ruskey~\cite{KanigFilliatre09wml} & 10 loc & 30 loc & $\sim$700 lop  \\ \hline
    \end{tabular}
\end{center}
  \caption{A summary of all programs proved so far using the \who\ tool}
  \label{fig:whoproofs}
\end{figure}

Fig.~\ref{fig:whoproofs} shows an overview over the programs we have been able
to prove using the \who\ tool. The first line summarizes the numbers of the
memo/ymemo example of the previous subsections.

The next two lines concern the two modules {\tt Array} and {\tt Hashtbl} of
the \ocaml\ standard library. For the {\tt Array} module, all functions
present in the \ocaml\ module have been proved, with the exception of the {\tt
Array.map} family, whose specification is very similar to {\tt iter}, and the
{\tt Array.sort} function, which is quite complicated to prove, but not very
interesting concerning higher-order functions or side effects. In this proof
development, arrays are modeled as references to functional arrays with the
usual axioms on access and update.

The proof development concerning the {\tt Hashtbl} module is ongoing work. The
difficult part of this development is the model of the \ocaml\ hash table; In
\ocaml, when one overwrites a previously existing entry in the table, the old
value still exists, and is only {\em shadowed} by the new key/value pair. If
one calls {\tt Hashtbl.remove} using this key, the old value reappears. Also,
if several bindings for a key exist, the key may be encountered several times,
during a call of {\tt Hashtbl.iter}, for example. A possible way of modeling
this is an association list instead of a simple mapping, where keys in front
of the list can shadow the same key further down in the list. 
% TODO Currently, only
%two functions of this module have been proved correct.

Finally, the last line shows the numbers for the proof development of the
Koda-Ruskey algorithm~\cite{KodaRuskey93}. A small, efficient
implementation~\cite{FilliatrePottier02} is shown to be correct. More details
about this development can be found in~\cite{KanigFilliatre09wml}.

\section{Related Work}
\label{sec:related}

%TODO rewrite
%TODO new citation for Ynot
Honda, Yoshida and Berger~\cite{Honda05anobservationally} present a
logic for imperative higher-order programs, using a slightly more
expressive language than ours. They present their calculus using rules
in Hoare logic style, while we presented a \wpre-calculus. Their
system does not provide the same modularity than ours, as one can not
reason modularly about side effects of functional arguments. It also
has the drawback of being difficult to implement, due to the presence
of a large number of non-structural rules. Our system, starting from a
correctly annotated program, is syntax-directed and thus entirely
automated. Their logic of proof obligations is non-standard, while we
obtain obligations in higher-order logic. However, their calculus can
deal easily with functions in references, even if the stored function
reads or modifies the reference it is stored in. While, with the
appropriate extensions, one can reason about functions in references
in our system, this particular kind of programs cannot be reasoned
about. Honda, Yoshida and Berger prove a completeness result for their
system. We have proved no such result. Their extended
system~\cite{BergerHondaYoshida05aliasing} has the same advantages and
drawbacks compared to ours, but additionally deals with aliasing
situations.

The Coq proof assistant~\cite{CoqManualV81}, in particular in
combination with the Program extension~\cite{sozeau07icfp}, can be
used to implement and prove arbitrary properties of purely functional
programs, using dependent types. The appeal of that approach is that
the only limit of expressiveness is the one of Coq. One can also do
meta-reasoning about programs inside Coq and, at the end of the
development, extract a certified implementation in \ocaml\ or
Haskell. The confidence in programs proved in this way is very high,
because in the end one obtains a Coq proof term that corresponds to
the correctness of the program. This term has been checked by Coq's
kernel. However, without extensions such as Ynot (see next paragraph),
it is impossible to implement (and reason about) effectful
computations. Another drawback of this approach is lack of automation.

The Ynot System~\cite{Nanevski08Awkward} is an extension to the Coq
proof assistant, capable of reasoning about imperative higher-order
programs, including effectful functions as arguments, using a monad in
which effectful computations can take place. It uses higher-order
separation logic as the device to speak about modifications of the
state. Ynot uses an (arguably less intuitive) monadic representation
of effectful computations. Ynot is more expressive than our system and
features modular reasoning (abstracting over the specification and
effect of a function as argument) while being able to reason about
aliasing situations thanks to separation logic.  On the other hand,
Ynot shares the Coq's main weakness: lack of automation. While we
haven't yet presented a platform to automatically discharge generated
proof obligations, this seems easily feasible, along the same vein as
Régis-Gianas and Pottier~\cite{regis-gianas-pottier-08}, who argue
that first-order SMT solvers~\cite{RanTin-SMTLIB} are suitable even
for higher-order proof obligations. They also present an encoding
scheme to support their claim. Their practical results are
promising. The Ynot system can not use this kind of automation because
it uses separation logic. Automated provers for separation logic
exist, but we are not aware of any attempts to use these tools in a
higher-order setting.

%TODO also say that an approach like Ynot ou Coq doesn't give any help
%for the mechanical part of the proof.

The Pangolin system, the implementation of the theoretical system
of~\cite{regis-gianas-pottier-08}, can also be used to reason about
purely functional programs. It is also one of the starting points of
our work, and our purely functional fragment is basically the Pangolin
system, although we removed algebraic data types for clarity of
presentation. For this reason, we share the potential advantages (good
automation, relatively simple system) and drawbacks (no
meta-reasoning or observational equality, no machine-checked proof
term). Pangolin cannot deal with side effects.

Another line of work consists in giving a translation from an
imperative to a functional language, with the aim of applying one of
the techniques for purely functional programs to the
translation. O'Hearn and Reynolds~\cite{ohearn-reynolds-00} present a
translation from Algol with first class references and dynamic
allocation to a linear $\lambda$-calculus. They impose an
alias-avoiding restriction which is similar to ours. Charguéraud and
Pottier~\cite{chargueraud-pottier-08} give a translation from Core ML
with first-class references, dynamic allocation and aliasing to a
functional language, using a very expressive type system with linear
capabilities. Both systems handle much more expressive languages than
ours. However, they do not describe a mechanism to obtain proof
obligations, for example a \wpre-calculus. Although we use a simpler
programming language, we present a complete cycle from imperative
programs to proof obligations. We also claim that extending our
language and our \wpre-calculus is simple, as long as aliasing remains
excluded.  We leave these extensions for future work.

Systems for verification of first-order programs like
Why~\cite{Filliatre00a} or Spec\#~\cite{BarnettLS04} are the other
starting points of this work. These systems generally deal very well
with usual features of first-order programs, for example arrays, use
SMT solvers to discharge proof obligations and strive for the best
possible automation. We try to improve on this work by adding
higher-order features, trying to maintain the degree of automation and
ease of use.
\section{Conclusion}
\label{sec:conclusion}

@@

We are currently investigating the possibility to obtain first-order proof
obligations, to be able to use SMT solvers~\cite{RanTin-SMTLIB}, which would
avoid manual proofs in many cases. An obvious solution would be to reuse
existing encodings of higher-order logic in first-order logic, such as the one
by Pottier and Gauthier~\cite{pottier-gauthier-hosc}. This technique has been
successfully employed by the Pangolin system~\cite{regis-gianas-pottier-08}.
Another option comes from our observation that in many cases, the proof
obligations for higher-order functions can actually be expressed in
first-order logic, when one replaces top-level {\em quantifications} of
functional arguments and invariants by top-level {\em declarations}. Doing
this translation manually, we have obtained first-order proof
obligations for the {\tt Array.iter} function which are all discharged by
SMT solvers. It is future work to find out if such a transformation can always
be carried out.

Currently, our effect system is quite limited. One possible extension is the
addition of subtyping: an effectful function $f$ of type $τ ->^ετ'$ can be seen
as a function of type $τ ->^{ε'}τ'$, if $ε\subseteq ε'$. In particular, such a
subtyping operation should come at no cost for program verification; it should
still come for free that $f$ does not modify the variables in $ε'\setminus ε$. It
would also be interesting to see if an ``impure'' function of type $τ
->^\emptyset τ'$ can be seen as pure function $τ -> τ'$ (and be used in the
logic).

Another possibility would be to modify the effect system itself to allow for
aliasing. In the setting with the aliasing restriction, our regions actually
become {\em singleton regions}, regions containing exactly one memory
location. Other, more powerful effect systems allow for {\em group regions},
with the possibility to temporarily {\em focus } on a particular memory cell
in that region. We see no fundamental difficulty in adopting such an effect
system, but haven't worked out the details.

Finally, from a programmer's point of view, it would be nice to have more
features in the programming language, such as recursive types and algebraic
datatypes. As long as so defined types verify a {\em positivity condition},
%TODO citation needed
these extensions would be completely orthogonal to the weakest precondition
calculus and the type and effect system used.
